Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2021-01-12 Nicolas.tex
YC comments included

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2020}, and assuming constant underlying stretch
    distribution for each age population made of Gaussian mixtures. We test our
    prediction against published samples \textbf{cut} to have marginal magnitude
    selection effects, so that any observed change is indeed of astrophysical
    and not observational origin. \textbf{In this first study, there are
    indications} that the underlying SN~Ia stretch distribution is evolving as a
    function of redshift, and that the young/old drifting model is a better
    description of the data than any time-constant model, including the
    sample-based asymmetric distributions usually used to correct Malmquist
    bias. The favored underlying stretch model is the bimodal one derived from
    \cite{rigault2020}: a high-stretch mode shared by both young and old
    environments, and a low-stretch mode exclusive to old environments. The
    precise impact of the redshift evolution of the SN Ia population intrinsic
    properties on cosmology remains to be studied. Yet, the astrophysical drift
    of the SN stretch distribution does affect current Malmquist bias
    corrections and thereby distances derived from SN affected by selection
effects. We highlight that such a bias will increase with surveys covering
increasingly larger redshift ranges, which is particularly important for LSST.}

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
sample, taking the farthest limits possible. But as this solution might be
over simplistic to create a complete sample, e.g., for it ignores follow-up
efficiency, we also consider another set of cuts to define a conservative
sample. This latter is smaller and therefore less statistically
constraining, but also inherently less prone to selection effects. If the
redshift drift is still significant in the conservative sample, it would be
even more significant in a carefully-tailored selection-free sample. The
goal of this study is to build the first implementation of a SN Ia
properties drift modeling before refining it on a more complex, yet more
accurate, follow-up analysis accounting at the same time for precise selection
modeling using, e.g., the SNANA package \citep{SNANA}.}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2020} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). \textbf{For this dataset, spectroscopic screening was
    done for candidates with $r \lesssim 19.5$. Then, redshift cuts were applied
    when selecting which SN Ia to follow, giving a redshift range of 0.02 < z <
    0.09 ; further reduced to $<0.08$ in \cite{rigault2020} for extracting local
host properties.} These 114 SNfactory SNe~Ia are thus in the volume-limited part
of the survey (Aldering et al., in prep.), and are therefore assumed to be a
random sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have a
large complete SN~Ia sample at $z<0.1$.

Finally, we include the HST sample from Pantheon \citep{strolger04}.
\textbf{These high-redshift SNe are of great interest as they provide the
    greatest leverage for testing evolution. While at these redshift the
    supernovae typing is challenging, the target classification was robust
enough to include them within the cosmological analysis \citep{scolnic2018a} and
we do not add further quality cuts. Section~\ref{sec:results} highlights that
our results are not dependent on the inclusion of this dataset.}

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \textbf{We observe
that the apparent fraction of low-stretch SNe (typically $x_1 < -1$) seems to
decrease as a function of redshift.} The evolution of the mean stretch is also
shown in Fig.~\ref{fig:modelall}, where the data are split in redshift bins of
regular sample size. We see that SNe~Ia at higher redshift have on average
larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower redshift
($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying stretch
distribution is \textbf{evolving with redshift}.

@@

\textbf{In section~\ref{ssec:cuts}, we have built a volume-limited sample from a
set of magnitude-limited ones, using a simple redshift limit. This
simplistic approach is statistically sub-optimal, but is enough to
demonstrate our key finding: a redshift evolution of stretch compatible with
the \cite{rigault2020} model. However, there remains the possibility that
complex selection function related to follow-up efficiencies below our fiducial
(or even conservative cuts) might significantly affect our sample, making it not
fully volume-limited, which could, in turn, bias our conclusion on the
astrophysical drift of the SNe~Ia population.}

\textbf{To test for the existence of potentially significant leftover selection
biases in our sample, we compare the stretch and color distributions of the
SNe~Ia in our sample between datasets of overlapping redshifts. If indeed we
have built a volume-limited sample, the distributions of any SNe~Ia
parameter at a given redshift should not depend on their survey of origin. When
considering a redshift range, this has to be narrow enough such that any drift
would be negligible.}

\textbf{The two samples that overlap the most in redshift are PS1 and SDSS in
the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus
the most likely to be affected by selection effects not eliminated by our
redshift cut (see the corresponding discussion in section~\ref{ssec:cuts}).
In that same redshift range, PS1 has 52 SNe~Ia, which are in the lowest
redshift bins and thus unlikely to have any selection issue. To identify
potential inconsistency between PS1 and SDSS, Fig.~\ref{fig:distrib}
presents the stretch and color distribution of both these surveys, together
with $p$-values of Kolmogorov-Smirnov (KS) tests. These $p$-values are too
high ($>10\%$) and do not indicate that these subsamples are inconsistent,
in agreement with the visual impression from Fig.~\ref{fig:distrib}.}

\textbf{We perform a similar analysis for PS1 and SNLS in the redshift range
$0.20 < z < 0.31$ where the same conclusion can be drawn, i.e., there is no
sign of discrepancy in the stretch and color distributions between the low-
and high-end of the SNLS and PS1 datasets, respectively. Nonetheless, the
small size of the SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1)
limits the sensitivity of the test, as only strong deviation would be
noticeable. Extending the redshift range to $0.20 < z < 0.40$ (though we
have no PS1 data above 0.3) enables to get 31 SNe~Ia from SNLS in total, yet
the stretch p-values remains high at 34\% showing no sign of inconsistency}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{\textbf{$x_1$ (left) and $c$ (right) distribution histograms of
            different surveys overlapping in redshift. \textit{Facing up}: SDSS
            and PS1 within the redshift range $0.10 < z < 0.20$; \textit{facing
            down}: PS1 and SNLS within the redshift range $0.20 < z < 0.31$.
            Error bars show the Poisson noise. Our stretch ``Base'' model is
            illustrated in orange at the mean redshift of the redshift ranges,
            $0.15$ and $0.25$, respectively. Kolmogorov-Smirnov test $p$-values
            are indicated on the top (resp. bottom) of each panel showing no
            sign that the SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$
            distributions are not drawn from the same underlying
    distributions.}}
    \label{fig:distrib}
\end{figure}

\textbf{We highlight that the SNe~Ia color is more prone to selection effects
than stretch as illustrated in Fig.~\ref{fig:maglim}; see also e.g., Fig.~3
of \cite{kessler2017}. Hence, since the color distribution shows no
significant hint of leftover selection effect, it further supports our claim
that our simple redshift-based selection criteria is sufficient to build the
SNe~Ia sample required to test the redshift evolution of the stretch
distribution.}

@@

\subsection{Base model applied to data}\label{sec:basemodelapplied}

@@

\textbf{In order to ensure that our results are not driven by the poorly-modeled
    HST sub-sample, we recompute $\Delta$AIC for each model excluding this
    dataset; we find (as listed in Table~\ref{tab:comp}) -3.4, -24.5, -27.2,
    -29.2, -26.9, -60.0 and -5.4, -17.4, -20.4, -22.4, -19.7, -57.9 for the
    fiducial and conservative samples, respectively. The consistency of these
    values with those in Table~\ref{tab:comp} show that the HST sub-sample does
not drive our conclusions.}

>>>>

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    evolution. In this paper, we study the dependence with redshift of the SN~Ia
    \texttt{SALT2.4} lightcurve stretch, a purely intrinsic SN property, to
    probe its potential redshift drift. The SN stretch has been shown to
    strongly correlate with the SN environment, notably with stellar age
    tracers. We model the underlying stretch distribution as a function of
    redshift, using the evolution of the fraction of young and old SNe~Ia as
    predicted by \cite{rigault2020}, and assuming constant underlying stretch
    distribution for each age population \textbf{consisting} of Gaussian
    mixtures. We test our prediction against published samples \textbf{cut} to
    have marginal magnitude selection effects, so that any observed change is
    indeed of astrophysical and not observational origin. \textbf{In this first
    study, there are indications} that the underlying SN~Ia stretch distribution
    is evolving as a function of redshift, and that the young/old drifting model
    is a better description of the data than any time-constant model, including
    the sample-based asymmetric distributions usually used to correct Malmquist
    bias. The favored underlying stretch model is the bimodal one derived from
    \cite{rigault2020}, \textbf{composed of} a high-stretch mode shared by both
    young and old environments, and a low-stretch mode exclusive to old
    environments. The precise impact of the redshift evolution of the SN Ia
    population intrinsic properties on cosmology remains to be studied. Yet, the
    astrophysical drift of the SN stretch distribution does affect current
    Malmquist bias corrections and thereby distances derived from SN affected by
selection effects. We highlight that such a bias will increase with surveys
covering increasingly larger redshift ranges, which is particularly important
for LSST.}

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
    sample, taking the limits as initially suggested by the previous
    completeness analysis. \textbf{However,} as this solution might be over
    simplistic to create a complete sample, e.g., for it ignores follow-up
    efficiency, we also consider another set of cuts to define a so-called
    conservative sample. This latter is smaller and therefore less statistically
    constraining, but also inherently less prone to selection effects. If the
    redshift drift is still significant in the conservative sample, it would be
    even more meaningful in a carefully-tailored selection-free sample. The goal
    of this study is to build the first implementation of a SN Ia properties
    drift modeling before refining it on a more complex, yet more accurate,
follow-up analysis also accounting for precise selection modeling using, e.g.,
the SNANA package \citep{SNANA}.}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2020} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). \textbf{For this dataset, spectroscopic screening was
    done for candidates with $r \lesssim 19.5$; redshift cuts were then applied
    when selecting which SN Ia to follow, resulting in a redshift range of $0.02
    < z < 0.09$, further reduced to $<0.08$ in \cite{rigault2020} for extracting
local host properties.} These 114 SNfactory SNe~Ia are thus in the
volume-limited part of the survey (Aldering et al., in prep.), and are therefore
assumed to be a random sampling of the underlying SN population. The SNfactory
sample is particularly useful for studying SN property drift, as it enables us
to have a large complete SN~Ia sample at $z<0.1$.

Finally, we include the HST sample from Pantheon \citep{strolger04}.
\textbf{These high-redshift SNe are of great interest as they provide the
    greatest leverage for testing evolution. While at these redshifts the
    supernovae typing is challenging, the target classification was robust
    enough to include them within the cosmological analysis \citep{scolnic2018a}
    and we do not add further quality cuts. Section~\ref{sec:results} highlights
    that, while compatible with it, our results are not dependent on the
    inclusion of this dataset.}

We present the stretch distribution and redshift histogram of these five surveys
up to their respective $z_{\lim}$ in Fig.~\ref{fig:sample}. \textbf{We observe
that the fraction of low-stretch SNe (typically $x_1 < -1$) seems to decrease as
a function of redshift; this is confirmed in Fig.~\ref{fig:modelall} in which}
the evolution of the mean stretch is shown, \textbf{with} the data split in
redshift bins of regular sample size. We see that SNe~Ia at higher redshift have
on average larger stretch ($0.34 \pm 0.10$ at $z\sim0.65$) than those at lower
redshift ($-0.17\pm 0.10$ at $z\sim0.05$), suggesting that the underlying
stretch distribution is \textbf{evolving with redshift}.

@@

\textbf{In section~\ref{ssec:cuts}, we have built volume-limited samples from a
    set of magnitude-limited ones, using simple redshift cuts. This simplistic
    approach is statistically sub-optimal, but should be enough to demonstrate
    our key finding: a redshift evolution of stretch compatible with
    \cite{rigault2020} model. However, the possibility remains that a complex
    selection function related to follow-up efficiencies below our fiducial (or
    even conservative) cuts could still affect our sample, making it not fully
volume-limited; this would, in turn, bias our conclusion on the astrophysical
drift of the SNe~Ia population. We now look at this possibility.}

\textbf{To test for the existence of potential leftover selection biases in our
    sample, we compare the stretch and color distributions of the SNe~Ia
    originating from different datasets over overlapping redshift ranges: these
    distributions should be similar. Note that the redshift range has to be
    narrow enough so that any drift would be negligible.}

\textbf{The two samples that overlap the most in redshift are PS1 and SDSS in
    the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
    corresponds to the 146 SNe~Ia from SDSS at the high redshift end and thus
    the most likely to be affected by residual selection effects (see the
    corresponding discussion in section~\ref{ssec:cuts}). Over that same
    redshift range, PS1 has 52 SNe~Ia, which are in the lowest redshift bins and
    thus unlikely to have any selection issue. To identify potential
    inconsistency between the PS1 and SDSS subsamples, Fig.~\ref{fig:distrib}
    (upper panels) compares the stretch and color distribution of both these
    surveys. The associated Kolmogorov-Smirnov (KS) similarity test $p$-values
are too high ($>10\%$) to conclude on any inconsistency, in agreement with the
visual impression from Fig.~\ref{fig:distrib}.}

\textbf{We perform a similar analysis for PS1 and SNLS over the redshift range
    $0.20 < z < 0.31$ (Fig.~\ref{fig:distrib}, lower panels), where the same
    conclusion can be drawn: there is no substantial sign of discrepancy in the
    stretch and color distributions between the low- and high-end of the SNLS
    and PS1 datasets, respectively. Nonetheless, the small size of the SNLS
    dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1) limits the sensitivity of
    the test, and only strong deviation would be noticeable. Extending the
redshift range to $0.20 < z < 0.40$ (though we have no PS1 data above 0.3)
allows to increase the SNLS subsample to 31, yet the stretch $p$-values remains
high (34\%) showing no sign of inconsistency.}

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{Article_figures/both-cut_SDSS_SNLS_PS1.pdf}
    \caption{\textbf{$x_1$ (left) and $c$ (right) distribution histograms of
            different surveys overlapping in redshift. \textit{Facing up}: SDSS
            and PS1 within the redshift range $0.10 < z < 0.20$; \textit{facing
            down}: PS1 and SNLS within the redshift range $0.20 < z < 0.31$.
            Error bars show the Poisson noise. Our stretch ``Base'' model is
            illustrated in orange at the mean redshift of the redshift ranges,
            $0.15$ and $0.25$, respectively. Kolmogorov-Smirnov test $p$-values
            are indicated on the top (resp. bottom) of each panel showing no
            sign that the SDSS and PS1 (resp. PS1 and SNLS) $x_1$ and $c$
            distributions are not drawn from the same underlying
    distributions}}
    \label{fig:distrib}
\end{figure}

\textbf{We finally highlight that the SNe~Ia color is more prone to selection
    effects than stretch as illustrated in Fig.~\ref{fig:maglim}; see also e.g.,
    Fig.~3 of \cite{kessler2017}. Hence, since the comparison of color
    distributions shows no significant hint of leftover selection effect, it
    further supports our claim that our simple redshift-based selection criteria
    is sufficient to build a complete SNe~Ia sample required to test the
    redshift evolution of the stretch distribution.}

@@

\subsection{Comparison to data}\label{sec:basemodelapplied}

@@

\textbf{In order to ensure that our results are not driven by the poorly-modeled
    HST sub-sample, we recompute $\Delta$AIC for each model excluding this
    dataset; we find that it does not change $\Delta$AIC by more than few
    tenths. The consistency of these values with those in Table~\ref{tab:comp}
show that the HST sub-sample does not drive our conclusions.}
